### 2014-11-04 thesis/abstract/abstract.tex
Lots of changes
Final version of thesis

Predicting the length of hospital stay for injured patients involves making a
judgement about how long they need to be hospitalised, so that doctors and
hospital staff can plan the appropriate course of treatment for the patient
to ensure they recover in the quickest possible time. This also ensures the
most efficient use of limited hospital resources. However, previous studies
have used a limited number of statistical techniques such as logistic
regression in order to predict the length of stay. Also, they have used only
manual feature selection and have not investigated the effect of automatic
feature selection on classifier performance.

Our work systematically evaluates and compares the performance of a number of
state-of-the-art classification algorithms  and feature selection methods,
and also assesses the effect of feature discretisation for all combinations
of classifiers and feature selectors.  Our evaluation were conducted on two
datasets, one from a trauma ward in a major Sydney hospital and another one
from a general hospital in Portugal. We propose a new classification algorithm
called Ranked Distance Nearest Neighbor that considers the
relative importance of each feature when determining the nearest neighbours.

For the trauma dataset, we were able to achieve an accuracy of 77.81\% with
1-nearest neighbour and an area under the curve of 0.846 with logistic
regression. This represents an improvement of 2.75\% and 0.034 respectively
over the previous best result on the same dataset. For the
general hospital dataset, the best results were 98.23\% accuracy with a
decision tree and 0.994 area under the curve with logistic regression,
support vector machine and K*. We also show that our proposed Ranked
Distance Nearest Neighbour improves accuracy and area under the curve over
traditional nearest neighbour algorithms. We comprehensively discuss
the results and suggest new directions for future studies.

>>>>

Predicting the length of hospital stay for injured patients involves making a
judgement about how long they need to be hospitalised, so that doctors and
hospital staff can plan the appropriate course of treatment for the patient
to ensure they recover in the quickest possible time. This also ensures the
most efficient use of limited hospital resources. However, previous studies
have only used a limited number of statistical techniques such as logistic
regression in order to predict the length of stay. Also, they have used only
manual feature selection and have not investigated the effect of automatic
feature selection on classifier performance.

Our work systematically evaluates and compares the performance of a number of
state-of-the-art classification algorithms and feature selection methods,
and also assesses the effect of feature discretisation for all combinations
of classifiers and feature selectors. Our evaluation were conducted on two
datasets, one from a trauma ward in a major Sydney hospital and another one
from a general hospital in Portugal. We propose a new classification algorithm
called Ranked Distance Nearest Neighbour that considers the
relative importance of each feature when determining the nearest neighbours.

For the trauma dataset, we were able to achieve an accuracy of 77.81\% with
1-nearest neighbour and an area under the receiver operating characteristic
curve of 0.846 with logistic
regression. This represents an improvement of 2.75\% and 0.034 respectively
over the previous best result on the same dataset. For the
general hospital dataset, the best results were 98.23\% accuracy with a
decision tree and 0.994 area under the curve with logistic regression,
support vector machine and K*. We also show that our proposed Ranked
Distance Nearest Neighbour improves accuracy and area under the curve over
traditional nearest neighbour algorithms. We comprehensively discuss
the results and suggest new directions for future studies.

### 2014-11-04 thesis/appendix/code.tex
Final version of thesis

In this appendix we display the source code of the most important scripts we
wrote in our to conduct our experiments. These are described in more detail
in Section \ref{sec:code}. To download the latest version of the code, please
visit \url{https://github.com/tianyupu/hons-thesis}.

\lstset{
  language = Python,
  texcl = false, % Parse comments as TeX
  backgroundcolor = \color{white},
  basicstyle = \footnotesize\ttfamily\singlespacing,
  commentstyle = \color{mygreen},
  keywordstyle = \color{blue},
  stringstyle = \color{mymauve},
  numberstyle = \tiny\color{mygray},
  numbers = left, % possible values are (none, left, right)
  numbersep = 5pt, % how far the line-numbers are from the code
  rulecolor = \color{black},
  stepnumber = 1, % the step between two line-numbers
  showstringspaces = false, % underline spaces within strings?
  literate = {~} {$\sim$}{1}, % make ~ nice
  literate = {_} {\textunderscore}{1}, % make _ nice
  breaklines = true,
  prebreak = \raisebox{0ex}[0ex][0ex]
  {\color{red}\ensuremath{\space\hookleftarrow}},
  postbreak = \raisebox{0ex}[0ex][0ex]
  {\color{red}\ensuremath{\hookrightarrow\space}},
  tabsize = 2
}

\section{Data Cleaning and Whitespace Removal}
\label{sec:python-preprocess}
\lstinputlisting{../code/sklearn/util.py}

@@

\subsection{Computing Confidence Intervals from Summarised Results}
\lstset{language=Python}
\lstinputlisting{../code/weka/compute_ci.py}

>>>>

In this appendix we display the source code of the most important programs we
wrote in our to conduct our experiments. These are described in more detail
in Section \ref{sec:code}. To download the latest version of the code, please
visit \url{https://github.com/tianyupu/hons-thesis}.

\lstset{
  language = Java,
  texcl = false, % Parse comments as TeX
  backgroundcolor = \color{white},
  basicstyle = \footnotesize\ttfamily\singlespacing,
  commentstyle = \color{mygreen},
  keywordstyle = \color{blue},
  stringstyle = \color{mymauve},
  numberstyle = \tiny\color{mygray},
  numbers = left, % possible values are (none, left, right)
  numbersep = 5pt, % how far the line-numbers are from the code
  rulecolor = \color{black},
  stepnumber = 1, % the step between two line-numbers
  showstringspaces = false, % underline spaces within strings?
  literate = {~} {$\sim$}{1}, % make ~ nice
  literate = {_} {\textunderscore}{1}, % make _ nice
  breaklines = true,
  prebreak = \raisebox{0ex}[0ex][0ex]
  {\color{red}\ensuremath{\space\hookleftarrow}},
  postbreak = \raisebox{0ex}[0ex][0ex]
  {\color{red}\ensuremath{\hookrightarrow\space}},
  tabsize = 2
}

\section{Data Cleaning and Whitespace Removal}
\lstset{language=Python}
\label{sec:python-preprocess}
\lstinputlisting{../code/sklearn/util.py}

@@

\subsection{Computing Confidence Intervals from Summarised Results}
\label{sec:compute-ci}
\lstset{language=Python}
\lstinputlisting{../code/weka/compute_ci.py}

### 2014-11-04 thesis/conclusion/conclusion.tex
Final version of thesis

In this thesis we conducted a systematic and empirical evaluation of a number
of classification algorithms and feature selection methods, as well as the
effect of discretisation, for predicting the length of hospital stay of trauma
patients. In particular, we introduced a new classification algorithm, Ranked
Distance Nearest Neighbours, and compared its accuracy and AUC with ten
other classifiers -- Zero-R, 1R, Na\"{i}ve Bayes, C4.5 decision tree, logistic
regression, support vector machine, $k$-nearest neighbour with $k=1$ and 20,
K* and multi-layer perceptron -- several of which (for example K*) have not
been used in existing work on LOS prediction. We also investigated a variety of
feature selection methods -- CFS, C4.5 wrapper, correlation coefficient,
information gain, 1R and expert selection -- encompassing both manual and
automatic methods, many of which have not been applied to the problem of LOS
prediction. Our experiments were conducted on two data sets, one from the
trauma ward of a Sydney hospital and the other a general hospital-wide data
set from Portugal.

@@

In our proposed Ranked Distance Nearest Neighbours algorithm, we only considered
one kind of weight assignment function, but it is not the only way to assign
these weights. There is room to investigate other types of functions, as well
as using weights obtained from another algorithm (such as logistic regression).
In addition to assigning the weights, we have also only investigated one method
of determining the relative importance of features: the correlation coefficient.
This leaves much room to use other ways of ranking features, such as the
information gain and 1R accuracy (which we have used in our work as feature
selectors).

>>>>

In this thesis we conducted a systematic and empirical evaluation of a number
of classification algorithms and feature selection methods, as well as the
effect of discretisation, for predicting the length of hospital stay of trauma
patients. In particular, we introduced a new classification algorithm, Ranked
Distance Nearest Neighbour, and compared its accuracy and AUC with ten
other classifiers -- Zero-R, 1R, Na\"{i}ve Bayes, C4.5 decision tree, logistic
regression, support vector machine, $k$-nearest neighbour with $k=1$ and 20,
K* and multi-layer perceptron -- several of which (for example K*) have not
been used in existing work on LOS prediction. We also investigated a variety of
feature selection methods -- CFS, C4.5 wrapper, correlation coefficient,
information gain, 1R and expert selection -- encompassing both manual and
automatic methods, many of which have not been applied to the problem of LOS
prediction. Our experiments were conducted on two data sets, one from the
trauma ward of a Sydney hospital and the other a general hospital-wide data
set from Portugal.

@@

In our proposed Ranked Distance Nearest Neighbour algorithm, we only considered
one kind of weight assignment function, but it is not the only way to assign
these weights. There is room to investigate other types of functions, as well
as using weights obtained from another algorithm (such as logistic regression).
In addition to assigning the weights, we have also only investigated one method
of determining the relative importance of features: the correlation coefficient.
This leaves much room to use other ways of ranking features, such as the
information gain and 1R accuracy (which we have used in our work as feature
selectors).

### 2014-11-04 thesis/introduction/introduction.tex
Final version of thesis

\begin{enumerate}
\item We systematically evaluated the performance of a catalogue of eleven
classification algorithms and five feature selection methods for predicting the
LOS of trauma patients. Three of the classifiers and four of the feature
selection methods have not been applied in LOS prediction to date, and only
one has been applied in our medical domain.

\item We assessed the effect of feature discretisation on classifier
performance. This has not been studied in previous work.

\item We proposed a new algorithm, Ranked Distance Nearest Neighbour, and
showed that it performed better than the standard nearest neighbour algorithm
in certain situations.

\item We conducted our evaluations on two independent data sets from different
medical domains: one from a trauma ward in a major Sydney hospital and the other
from a general hospital in Portugal. The collection, processing and
understanding of the data involved collaboration with medical experts (Dr
Michael Dinh from the Royal Prince Alfred Hospital in Sydney) and researchers
from Portugal (Associate Professor Paulo Cortez from the University of Minho).
Not only is the evaluation of a second data set in one work uncommon in the
literature, the inclusion of more than one data set allowed us to draw
additional insights about the classifiers and feature selections we used
which would not have been possible with only one set of data.

@@

Our findings are presented and discussed in Chapter \ref{chap:results}.
We conclude in Chapter \ref{chap:conclusion}, with indications for future
work in this area.

>>>>

\begin{enumerate}
\item We systematically evaluated the performance of a catalogue of eleven
classification algorithms and five feature selection methods for predicting the
LOS of trauma patients. Three of the classifiers and four of the feature
selection methods have not been applied in LOS prediction to date, and only
one has been applied in our medical domain;

\item We assessed the effect of feature discretisation on classifier
performance. This has not been studied in previous work;

\item We proposed a new algorithm, Ranked Distance Nearest Neighbour, and
showed that it performed better than the standard nearest neighbour algorithm
in certain situations;

\item We conducted our evaluations on two independent data sets from different
medical domains: one from a trauma ward in a major Sydney hospital and the other
from a general hospital in Portugal. The collection, processing and
understanding of the data involved collaboration with medical experts (Dr
Michael Dinh from the Royal Prince Alfred Hospital in Sydney) and researchers
from Portugal (Associate Professor Paulo Cortez from the University of Minho).
Not only is the evaluation of a second data set in one work uncommon in the
literature, its inclusion enabled us to draw
additional insights about the classifiers and feature selections we used
which would not have been possible with only one set of data;

@@

Our findings are presented and discussed in Chapter \ref{chap:results}.
We conclude in Chapter \ref{chap:conclusion}, with indications for possible
avenues of future research.

### 2014-11-04 thesis/litreview/litreview.tex
Final version of thesis

\section{Introduction}
Hospitals often have limited funding, and therefore managing and utilising
their resources efficiently has always been a priority \cite{Walczak2003}.
One key measurement of hospital activity, health care utilisation, and hospital
cost is the patient length of stay \citep{Omachonu2004,Ng2006}. The length of
stay has been used in a wide variety of medical domains, such as burns
 \citep{Yang2010}, intensive
care \citep{Tu1993,Harper2005,Perez2006,Dybowski1996},
psychiatry \citep{Lowell1997}, and acute pancreatitis \citep{Pofahl1998}, in
order to assist hospitals in allocating beds and other patient management
resources as efficiently as possible. However, given the many different
conditions that hospitals manage, the length of stay of a patient is
not easy to predict \citep{Walczak2003}, especially upon a patient's admission.
Many methods have been explored in the literature, from the realm of statistics
to techniques in data mining, and they have been explored for many different
medical domains.

One of the more well-understood and widely used statistical methods for
predicting the length of stay is \textit{logistic regression} \citep{Tu1996}.
Logistic regression models are straightforward to construct and use via the
help of statistical software packages such as
SAS\footnote{SAS Institute, Cary, NC, USA}, and
also make the contribution of each independent variable to the final
prediction explicit. Thus they is easy to interpret when compared to more
``black box'' approaches such as artificial neural networks \citep{Adams2012}.
A wide variety of other statistical approaches have been proposed, notably:
various scoring systems from univariate analysis \citep{Adams2012,Lavoie2005},
linear regression \citep{Yang2010}, exponential models \citep{Clark2007},
survival analysis \citep{Vasilakis2005}, and Markov
models \citep{Perez2006,Jain1989,Kapadia2000}.

Out of the development of the various statistically-motivated models, data
mining techniques, particularly artificial neural networks,
began to become more widely adopted following a number of
early applications to length of stay prediction in intensive
care \citep{Tu1993,Dybowski1996}, in addition to disease diagnosis and mortality
prediction \citep{Silva2006}. Artificial neural networks are able to detect
complex and non-linear relationships between input and output variables and can
be developed using many different techniques. Much work has been done in using
neural networks to predict patient length of stay in various medical conditions
such as acute pancreatitis \citep{Walczak2003}, gastroenteritis \citep{Ng2006},
surgical intensive care \citep{Buchman1994,Tu1993} and
psychiatry \citep{Lowell1997}. Other data mining techniques have also been
applied, such as support vector machines and tree-based
methods \citep{Harper2005}.

Given the benefits to the hospital of accurate length of stay prediction,
this research aims to build upon improving length of stay prediction for trauma
patients by utilising state-of-the-art data mining techniques. Being able to
predict the length of stay for trauma patients is especially crucial because of
the scarcity of patient care resource in this area \citep{Yang2010}.
Additionally,
despite the wealth of research into length of stay prediction in various
medical domains, there has been little work done in generalising models from
one domain to another, or to predict a combination of outcomes (not only
length of stay, but mortality, or rehabilitation requirement). Being able to
do this would mean that hospital staff are better equipped to handle patient
cases and prepare for possible critical outcomes well in advance, resulting
in time and resources saved for both hospital and patient.

@@

This is the method that Gabbe et al. used in their study to identify factors
that contributed to trauma patient length of stay and use these factors to
subsequently predict length of stay \citep{Gabbe2005}. The prediction problem
they proposed was essentially a binary classification problem, to which
logistic regression is well suited \citep{Tu1996}: will length of stay be
\textit{less than 5 days}, or \textit{equal to 5 days}? They used a
\textit{backward} stepwise selection algorithm, whereby all available variables
were initially included in the model and then progressively eliminated if their
likelihood ratio (LR) test $P$-value was $>0.15$. Features could only be
re-included if $P<0.05$. For length of stay
prediction, this left them with 9 features out of the 15 that they originally
included. To validate their model, they used a data set collected from an
independent source which had slightly different statistical characteristics for
the various features and outcomes. Their length of stay prediction model was
found to have 95\% sensitivity and 24\% specificity, with an area under the ROC
curve (AUC) of 0.62. This indicates that the model is not able to accurately
discriminate
length of stay cases of less than 5 days and equal to 5 days (i.e., the two
desired outcomes) very well, performing only
slightly better than simply random guessing (which has AUC 0.50).

@@

\section{Data Mining Methods}
Data mining is the analysis of (often large) observational data sets to find
unsuspected relationships and to summarise the data in novel ways that are both
understandable and useful to the data owner \citep{Hand2001}. Compared to
traditional statistical techniques, it is a relatively recent field, having
emerged in the 1990s, and has
given researchers a variety of new tools to analyse and understand biomedical
datasets \citep{Bellazzi2008,Yoo2012}. Bellazzi \& Zupan \citep{Bellazzi2008}
have written a
comprehensive review paper of data mining techniques in clinical prediction
settings, and we refer the reader to their article for a more detailed
treatment of the techniques and their applications. Here we will describe the
major methods in the context of their applicability to the problem of
predicting length of hospital stay.

\subsection{Artificial Neural Networks}
Artificial neural networks have been widely used in clinical prediction
following their use in diagnosing acute myocardial infarction and various
applications in radiology \citep{Baxt1995}.
One of the earliest applications of neural networks to
length of stay prediction was by Tu \& Guerriere, who used a neural network
with the following structure \citep{Tu1993} to predict length of stay in
intensive care following cardiac surgery:
\begin{figure}[h]
\centering
\includegraphics[scale=0.95]{images/a2-nn}
\caption{A graphical representation of a neural network with one hidden layer,
used by Tu \& Guerriere in predicting LOS.}
\end{figure}

@@

Using this technique,
Tu \& Guerriere wanted to predict whether a patient was going to stay
\textit{less than 2 days} or \textit{more than 2 days} following cardiac
surgery. After testing their trained neural network on the test data, they
found that it had an AUC of 0.6960, indicating a reasonable amount of
discriminating ability between the two outcomes. Even though we cannot compare
their result to any of the studies conducted on length of stay prediction
using logistic regression techniques, there are a number of well-known reasons
why neural networks are advantageous to modelling complex clinical situations:
\begin{enumerate}
\item Neural networks are able to model arbitrarily complex relations between
independent and dependent variables \citep{Tu1996,Buchman1994}, especially if
the relation is of a non-linear nature
\item Neural networks are able to handle models where the features (independent
variables) are not completely independent of each other
\item Neural networks can be ``trained'' in different ways, which changes their
predictive power -- this has been studied by Walczak et al. \citep{Walczak2003},
discussed in the next section
\end{enumerate}

However, there were a few difficulties in using the neural network as Tu \&
Guerriere discovered, which are problems often cited in the use of such
networks. These include a high sensitivity to the starting parameters (such as
network architecture\footnote{For example, how many hidden layers to have, and
how many nodes in each hidden layer. The difficulty of finding the right
architecture is pointed out by Tu \& Guerriere themselves -- in their paper
they state, ``The optimal structure of the neural network model was determined
by `trial and error' ''.}
and learning rate\footnote{A constant that is set by the
network developer that governs how much the connection weights should adjust at
each iteration of the training algorithm.}), as well as the difficulty in
inducing sensible rules from the model (even by domain
experts) \citep{Bellazzi2008}. Additionally, they are also prone to overfitting
and require substantially more computational power to train when compared to
the standard logistic regression technique, and do not always perform
better \citep{Harper2005}. A number of studies have attempted to address some of
these issues in length of stay prediction, building on from Tu \& Guerriere's
work, and these will be described in the next section.

@@

Dybowski et al. decided to address the problem of neural network architecture
by applying a genetic algorithm approach to selecting the best
structure \citep{Dybowski1996} --
an improvement from the ``trial and error'' method Tu \& Guerriere used. It is
interesting to briefly discuss their method even though they were interested
in predicting the outcome of critically ill patients in intensive care, rather
than the length of stay which is the focus of our work. The genetic algorithm
in choosing an optimal neural network structure is as
follows \citep{Dybowski1996}:
\begin{enumerate}
\item A large population of random neural networks is created, each with
a random number of nodes in three hidden layers.
\item Each neural network optimises
its connection weights with respect to the same data set using
backpropagation.
\item Each neural network is expressed as a string which represents the
substructure of the network -- this is the ``chromosome''.
\item The ``chromosomes'' are copied to form a new population whereby the
expected relative frequency of a chromosome is proportional to how well
its neural network performed -- this is the ``survival of the fittest''
aspect.
\item The replica chromosomes undergo crossover so that better-performing
components can be combined. Mutations are also present, through the random
modification of a chromosome.
\item The process is repeated until a best-performing neural network forms.
\end{enumerate}
Through this Dybowski et al. were able to show that successive ``generations''
of neural networks had greater AUC measures, indicating better predictive
ability. Additionally, after seven generations, they obtained a neural network
that had an AUC of 0.863 compared to 0.753 for an equivalent logistic
regression model.

@@

Additionally, Walczak et al. also examined not only backpropagation, which
is the training algorithm used by most of the applications, but also radial
basis function and fuzzy ARTMAP. Using the mean absolute difference, they
found that a single hidden layer backpropagation neural network performed
better than all other networks in pediatric trauma, and was able to predict
the length of stay to within 1 day for 52\% of all cases. They also obtained
promising results using the fuzzy ARTMAP method, which warrants further
investigation as this training method has not been applied to modelling
length of stay in other medical domains.

@@

CART has been extensively used in medical decision making, and at its heart
involves an algorithm that is used to split the data set into sub-groups of
increasing similarity. At each node in the tree, the algorithm examines the
independent variables and splits the data set on the variable that produces
a node with the smallest variance (i.e., the greatest similarity). This is
continued until defined stopping rules are met. One of the major reasons that
such tree-based methods are more popular than neural networks in the
healthcare domain is that the rules it ``learns'' are very simple to extract
from the resulting tree, unlike the ``black box'' that neural networks commonly
are. Harper found that CART performed almost as well as neural networks in
predicting intensive care length of stay, and much better in predicting
hospital length of stay in terms of predicting accuracy (the only evaluation
statistic reported). He also found that CART was superior to regression, a
finding that Yang et al. also reported when predicting the length of stay of
burn patients \citep{Yang2010}.

Yang et al. investigated the performance of linear regression, M5 (another
tree-based classification method), and support vector machines in predicting
length of stay for burn patients. The difference between their tree method,
M5, with that of Harper (CART), is that the nodes contain linear regression
functions instead of the usual discrete (two-class) outputs that are used to
successively split the data set. Additionally, they also use support vector
machines (SVMs), perhaps the most powerful classification algorithm in terms of
predictive accuracy \citep{Bellazzi2008}. SVMs work on the principle that when
the data is represented in a high enough dimensional space, a linear hyperplane
can be found where the margin between two different records is as large as
possible. In Figure \ref{fig:svm-example} from Yoo et al. \citep{Yoo2012},
hyperplane 2 is the better plane because its margin is as large as possible. 
\begin{figure}[h]
\centering
\includegraphics[scale=1.05]{images/a2-svm}
\caption{An example of finding the largest margin hyperplane for a two-class
support vector machine classifier.}
\label{fig:svm-example}
\end{figure}

@@

\section{Summary}
Despite the wide range of statistical and data mining techniques that have been
used to predict the hospital length of stay, we have raised a few avenues for
further work throughout the literature review. Firstly, there has been no work
done to predict the length of stay of trauma patients using data mining
techniques, which is something worth investigating as trauma is a complex
medical condition. We have pointed out that there is also value in
investigating a number of the best data mining techniques, i.e. using more
than just neural networks. There is also value in investigating the performance
of such a model on data sets from different locations, since most studies are
only validated on one or two data sets. Additionally, length of stay prediction
has largely been restricted to specific medical domains, and therefore it is
worthwhile to examine if there are factors present in other domains which allow
for a more general model of the length of stay.

>>>>

\section{Introduction}
Hospitals often have limited funding, and therefore managing and utilising
their resources efficiently has always been a priority \cite{Walczak2003}.
One key measurement of hospital activity, health care utilisation, and hospital
cost is the patient length of stay \citep{Ng2006,Omachonu2004}. The length of
stay has been used in a wide variety of medical domains, such as burns
 \citep{Yang2010}, intensive
care \citep{Dybowski1996,Harper2005,Perez2006,Tu1993},
psychiatry \citep{Lowell1997}, and acute pancreatitis \citep{Pofahl1998}, in
order to assist hospitals in allocating beds and other patient management
resources as efficiently as possible. However, given the many different
conditions that hospitals manage, the length of stay of a patient is
not easy to predict \citep{Walczak2003}, especially upon a patient's admission.
Many methods have been explored in the literature, from the realm of statistics
to techniques in data mining, and they have been explored for many different
medical domains.

One of the more well-understood and widely used statistical methods for
predicting the length of stay is \textit{logistic regression} \citep{Tu1996}.
Logistic regression models are straightforward to construct and use via the
help of statistical software packages such as
SAS\footnote{SAS Institute, Cary, NC, USA}, and
also make the contribution of each independent variable to the final
prediction explicit. Thus they are easier to interpret when compared to more
``black box'' approaches such as artificial neural networks \citep{Adams2012}.
A wide variety of other statistical approaches have been proposed, notably:
various scoring systems from univariate analysis \citep{Adams2012,Lavoie2005},
linear regression \citep{Yang2010}, exponential models \citep{Clark2007},
survival analysis \citep{Vasilakis2005}, and Markov
models \citep{Jain1989,Kapadia2000,Perez2006}.

Out of the development of the various statistically-motivated models, data
mining techniques, particularly artificial neural networks,
began to become more widely adopted following a number of
early applications to length of stay prediction in intensive
care \citep{Dybowski1996,Tu1993}, in addition to disease diagnosis and mortality
prediction \citep{Silva2006}. Artificial neural networks are able to detect
complex and non-linear relationships between input and output variables and can
be developed using many different techniques. Much work has been done in using
neural networks to predict patient length of stay in various medical conditions
such as acute pancreatitis \citep{Walczak2003}, gastroenteritis \citep{Ng2006},
surgical intensive care \citep{Buchman1994,Tu1993} and
psychiatry \citep{Lowell1997}. Other data mining techniques have also been
applied, such as support vector machines and tree-based
methods \citep{Harper2005}.

Given the benefits to the hospital of accurate length of stay prediction,
this research aims to build upon improving length of stay prediction for trauma
patients by utilising state-of-the-art data mining techniques. Being able to
predict the length of stay for trauma patients is especially crucial because of
the scarcity of patient care resource in this area \citep{Yang2010}.
Additionally,
despite the wealth of research into length of stay prediction in various
medical domains, there has been little work done in generalising models from
one domain to another, or to predict a combination of outcomes (not only
length of stay, but mortality or rehabilitation requirement). Being able to
do this would mean that hospital staff are better equipped to handle patient
cases and prepare for possible critical outcomes well in advance, resulting
in time and resources saved for both hospital and patient.

@@

This is the method that Gabbe et al. used in their study to identify factors
that contributed to trauma patient length of stay and use these factors to
subsequently predict length of stay \citep{Gabbe2005}. The prediction problem
they proposed was essentially a binary classification problem, to which
logistic regression is well suited \citep{Tu1996}: will length of stay be
\textit{less than 5 days}, or \textit{equal to 5 days}? They used a
\textit{backward} stepwise selection algorithm, whereby all available variables
were initially included in the model and then progressively eliminated if their
likelihood ratio (LR) test $P$-value was greater than 0.15. Features could
only be re-included if $P<0.05$. For length of stay
prediction, this left them with 9 features out of the 15 that they originally
included. To validate their model, they used a data set collected from an
independent source which had slightly different statistical characteristics for
the various features and outcomes. Their length of stay prediction model was
found to have 95\% sensitivity and 24\% specificity, with an area under the ROC
curve (AUC) of 0.62. This indicates that the model is not able to accurately
discriminate
length of stay cases of less than 5 days and equal to 5 days (i.e., the two
desired outcomes) very well, performing only
slightly better than simply random guessing (which has AUC 0.50).

@@

\section{Data Mining Methods}
Data mining is the analysis of (often large) observational data sets to find
unsuspected relationships and to summarise the data in novel ways that are both
understandable and useful to the data owner \citep{Hand2001}. Compared to
traditional statistical techniques, it is a relatively recent field, having
emerged in the 1990s, and has
given researchers a variety of new tools to analyse and understand biomedical
datasets \citep{Bellazzi2008,Yoo2012}. Bellazzi and Zupan \citep{Bellazzi2008}
have written a
comprehensive review paper of data mining techniques in clinical prediction
settings, and we refer readers to their article for a more detailed
treatment of the techniques and their applications. Here we will describe the
major methods in the context of their applicability to the problem of
predicting length of hospital stay.

\subsection{Artificial Neural Networks}
Artificial neural networks have been widely used in clinical prediction
following their use in diagnosing acute myocardial infarction and various
applications in radiology \citep{Baxt1995}.
One of the earliest applications of neural networks to
length of stay prediction was by Tu and Guerriere, who used a neural network
with the following structure \citep{Tu1993} to predict length of stay in
intensive care following cardiac surgery:
\begin{figure}[h]
\centering
\includegraphics[scale=0.95]{images/a2-nn}
\caption{A graphical representation of a neural network with one hidden layer,
used by Tu \& Guerriere in predicting LOS.}
\end{figure}

@@

Using this technique,
Tu \& Guerriere wanted to predict whether a patient was going to stay
\textit{less than 2 days} or \textit{more than 2 days} following cardiac
surgery. After testing their trained neural network on the test data, they
found that it had an AUC of 0.6960, indicating a reasonable amount of
discriminating ability between the two outcomes. Even though we cannot compare
their result to any of the studies conducted on length of stay prediction
using logistic regression techniques, there are a number of well-known reasons
why neural networks are advantageous to modelling complex clinical situations:
\begin{enumerate}
\item Neural networks are able to model arbitrarily complex relations between
independent and dependent variables \citep{Buchman1994,Tu1996}, especially if
the relation is of a non-linear nature
\item Neural networks are able to handle models where the features (independent
variables) are not completely independent of each other \cite{Tu1996}
\item Neural networks can be ``trained'' in different ways, which changes their
predictive power -- this has been studied by Walczak et al. \citep{Walczak2003},
discussed in the next section
\end{enumerate}

However, there were a few difficulties in using the neural network as Tu \&
Guerriere discovered, which are problems often cited in the use of such
networks. These include a high sensitivity to the starting parameters (such as
network architecture\footnote{For example, how many hidden layers to have, and
how many nodes in each hidden layer. The difficulty of finding the right
architecture is pointed out by Tu \& Guerriere themselves -- in their paper
they state, ``The optimal structure of the neural network model was determined
by `trial and error' ''.}
and learning rate\footnote{A constant set by the
network developer that governs how much the connection weights should adjust at
each iteration of the training algorithm.}), as well as difficulty in
inducing sensible rules from the model (even by domain
experts) \citep{Bellazzi2008}. Additionally, they are also prone to overfitting
and require substantially more computational power to train when compared to
the standard logistic regression technique, and do not always perform
better \citep{Harper2005}. A number of studies have attempted to address some of
these issues in length of stay prediction, building on from Tu \& Guerriere's
work, and these will be described in the next section.

@@

Dybowski et al. decided to address the problem of neural network architecture
by applying a genetic algorithm approach to selecting the best
structure \citep{Dybowski1996} --
an improvement from the ``trial and error'' method Tu \& Guerriere used. It is
interesting to briefly discuss their method even though their focus was
in predicting the outcome of critically ill patients in intensive care, rather
than the length of stay which is the focus of our work. The genetic algorithm
in choosing an optimal neural network structure is as
follows \citep{Dybowski1996}:
\begin{enumerate}
\item A large population of random neural networks is created, each with
a random number of nodes in three hidden layers.
\item Each neural network optimises
its connection weights with respect to the same data set using
backpropagation.
\item Each neural network is expressed as a string which represents the
substructure of the network -- this is the ``chromosome''.
\item The ``chromosomes'' are copied to form a new population whereby the
expected relative frequency of a chromosome is proportional to how well
its neural network performed -- this is the ``survival of the fittest''
aspect.
\item The replica chromosomes undergo crossover so that better-performing
components can be combined. Mutations are also present, through the random
modification of a chromosome.
\item The process is repeated until a best-performing neural network forms.
\end{enumerate}
Through this Dybowski et al. were able to show that successive ``generations''
of neural networks had greater AUC measures, indicating better predictive
ability. Additionally, after seven generations, they obtained a neural network
that had an AUC of 0.863 compared to 0.753 for an equivalent logistic
regression model.

@@

Additionally, Walczak et al. examined not only backpropagation, which
is the training algorithm used by most of the applications, but also radial
basis function and fuzzy ARTMAP. Using the mean absolute difference, they
found that a single hidden layer backpropagation neural network performed
better than all other networks in pediatric trauma, and was able to predict
the length of stay to within 1 day for 52\% of all cases. They also obtained
promising results using the fuzzy ARTMAP method, which warrants further
investigation as this training method has not been applied to modelling
length of stay in other medical domains.

@@

CART has been extensively used in medical decision making, and at its heart
involves an algorithm that is used to split the data set into sub-groups of
increasing similarity. At each node in the tree, the algorithm examines the
independent variables and splits the data set on the variable that produces
a node with the smallest variance (i.e., the greatest similarity). This is
continued until defined stopping rules are met. One of the major reasons that
such tree-based methods are more popular than neural networks in the
healthcare domain is that the rules it ``learns'' are very simple to extract
from the resulting tree, unlike the ``black box'' nature of neural network
predictions.
Harper found that CART performed almost as well as neural networks in
predicting intensive care length of stay, and much better in predicting
hospital length of stay in terms of predicting accuracy (the only evaluation
statistic reported). He also found that CART was superior to regression, a
finding that Yang et al. also reported when predicting the length of stay of
burn patients \citep{Yang2010}.

Yang et al. investigated the performance of linear regression, M5 (another
tree-based classification method), and support vector machines in predicting
length of stay for burn patients. The difference between their tree method,
M5, with that of Harper (CART), is that the nodes contain linear regression
functions instead of the usual discrete (two-class) outputs that are used to
successively split the data set. Additionally, they also use support vector
machines (SVMs), perhaps the most powerful classification algorithm in terms of
predictive accuracy \citep{Bellazzi2008}. SVMs work on the principle that when
the data is represented in a high enough dimensional space, a linear hyperplane
can be found where the margin between two different records is as large as
possible. In Figure \ref{fig:svm-example} from Yoo et al. \citep{Yoo2012},
hyperplane 2 is the better plane because its margin is as large as possible,
meaning that the SVM has the best generalising ability to classify unseen
examples.
\begin{figure}[h]
\centering
\includegraphics[scale=1.05]{images/a2-svm}
\caption{An example of finding the largest margin hyperplane for a two-class
support vector machine classifier.}
\label{fig:svm-example}
\end{figure}

@@

\section{Summary}
Despite the wide range of statistical and data mining techniques that have been
used to predict the hospital length of stay, we have raised a few avenues for
further work throughout the literature review. Firstly, there has been no work
done to predict the length of stay of trauma patients using data mining
techniques, which is something worth investigating as trauma is a complex
medical condition. We have pointed out that there is also value in
investigating a number of the best data mining techniques, i.e. using more
than just neural networks. An investigation of the performance
of such a model on data sets from different locations is also warranted, since
most studies are
only validated on one or two data sets. Additionally, length of stay prediction
has largely been restricted to specific medical domains, and therefore it is
worthwhile to examine if there are factors present in other domains which allow
for a more general model of the length of stay.
